---
title: "Class 5: Data Viz with ggplot2"
author: "Samarah Zaman (PID: A17411141)"
format: pdf
toc: true
---

## Background

There are many graphics systems in R for making plots and figures. These include so-called *"base R" graphics* like the `plot()` function and add on packages like **ggplot2**.

Let's compare how we make a simple figure with these two systems:

We can use the in-built `cars` dataset:

```{r}
head(cars)
```

```{r}
plot(cars)
```

Before I can use ggplot2 I need to install it on my computer.
To do this we can use the function `install.packages("ggplot2")`

>**N.B.** We never run `install.packages()` in our quorto doc (we run it once only in our R console) as it would re-install the package every time we render our quorto

Once installed we need to load up the pagckage into our R brain:

```{r}
library(ggplot2)
```

The main function in ggplot2 package is called `ggplot()`

```{r}
ggplot(cars)
```

Every ggplot has at least 3 layers :

- the **data** (a data.frame of the stuff we want to plot)
- the **aes**thetics (how the data maps to the plot)
- the **geom** layer (how you want the plot drawn, e.g. points, lines, etc. )

```{r}
ggplot(cars)+
  aes(x=speed, y=dist) +
  geom_point()
```

## Add some custom features

Let's add a trend line that shows the relationship betweeen speed and distance.

```{r}
ggplot(cars)+
  aes(x=speed, y=dist) +
  geom_point()+
  geom_smooth()+
  theme_bw() + 
  labs(title= "Stoping Dist of old cars",
       x="Speed (MPH)",
       y="Distance (ft)")
```


Q. Can you make `geom_smooth()` function produce a linear straight line fit to the data and turn off the "gray" error region.

```{r}
ggplot(cars)+
  aes(x=speed, y=dist) +
  geom_point()+
  geom_smooth(method=lm, se = FALSE)+
  theme_bw() + 
  labs(title= "Stoping Dist of old cars",
       x="Speed (MPH)",
       y="Distance (ft)")
```

## Gene expression figures

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

```{r}
sum(genes$State == "up")
```

A useful new function in this context is the `table()` function: 

```{r}
table(genes$State)
```

My first plot attempt

```{r}
ggplot(genes)+ 
  aes(Condition1, Condition2, col=State)+ 
  geom_point()+
  scale_colour_manual( values=c("blue","gray","red") )+
    labs(title="Gene Expresion Changes Upon GLP-1 Inhibitor Treatment",
         x="Control (No drug) ",
         y="Drug Treatment")
```

## Going further

Here we read the famous gapminder dataset :

```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
head(gapminder)
```

>Q. How many entries (i.e. rows) are in this dataset?

```{r}
nrow(gapminder)
```

>Q.  How many different "country" entries are in this dataset?

```{r}
length(table(gapminder$country))
```
```{r}
length(unique(gapminder$country))
```
Let's make our first plot of the entire dataset:

Plot of "gdpPercap" vs "lifeExp" colored by "continent"
```{r}
p <- ggplot(gapminder)+ 
  aes(gdpPercap, lifeExp, col=continent)+ 
  geom_point(alpha=0.3)
```

```{r}
p
```
I can add more layers to `p`
```{r}
p+
  facet_wrap(~continent)
```

Make a plot for years 1977 and 2007 only (not all the years in the dataset).


>Q.First use **dplyr** package and the `filter()` function from that package to extract the rows from the year 2007.

```{r}
library(dplyr)
```
```{r}
g07<- filter(gapminder, year == 2007) 
g77<- filter(gapminder, year == 1977) 
g <- filter(gapminder, year == 2007 | year == 1977) 
```

```{r}
ggplot(g)+ 
  aes(gdpPercap, lifeExp, col=continent, size=pop)+ 
  geom_point()+
  facet_wrap(~year)
```

>Q. Make a histogram of lifeExp colored by continent

```{r}
ggplot(gapminder)+
  aes(lifeExp, fill=continent)+
  geom_histogram() 
```

>Q. Make a histogram of lifeExp faceted by continent

```{r}
ggplot(gapminder)+
  aes(lifeExp, color=continent)+
  geom_histogram() +facet_wrap(~continent)
```


